Lens Modeling in \sw

### 2014-06-16 text/partA.tex
editing

Since that early days, more than 400 secure lenses have been found
(see Figure \ref{fig:masterlens}) and detailed modelling of the mass
distribution is part of any research using multiply-imaging lenses.
But so far no modelling study has used anything approaching 400
lenses.  The largest single one \citep{2009ApJ...703L..51K} used 58
lenses to uncover the distribution of dark matter around galaxies.
\cite{2011ApJ...740...97L} combined lens models of 21 galaxies with
models of their stellar populations, to infer the relation between
stars and dark matter in galaxies.  \cite{2014MNRAS.437..600S}
modelled 18 time-delay lenses together to infer cosmological
parameters.

\begin{figure}
\centering
\includegraphics[width=\hsize]{fig/lenssky}
\caption{Sky distribution of 423 published lenses considered secure
  (from the Masterlens catalog at the University of Utah, maintained
  by Joel Brownstein and Leonidas Moustakas.) The map is in
  Hammer-Aitoff projection, with North up and $\rm RA=0$ in the
  middle. The empty swathes to the left and right of center are the
  Milky Way. {\em Should we drop this figure?  It's not relevant to
    modelling, but as there seems to be no publicly available summary
    of all known lenses, it does help set context.}}
\label{fig:masterlens}
\end{figure}

Integrating modelling into the lens discovery process would be a great
advantage.  In this paper, we develop a technique towards this goal,
working with the \sw lens search (Marshall et al, in prep, More et al,
in prep).  \sw is a citizen science project\footnote{\tt
  http://www.spacewarps.org} in which volunteers are presented
sky-survey images and invited to identify lens candidates.  Simulated
lenses are mixed in with the data, both to help train volunteers on
what to look for, and to estimate completeness and purity.  The
motivation for \sw is to enable volunteers, some of whom had
previously serendipitously identified lens candidates on earlier
citizen-science surveys, to make discoveries missed in automatic
searches by software robots.  Robots are very good with clean lensing
system in uncrowded fields with high signal-to-noise, but in general
test situations \citep{2009ApJ...694..924M} robots miss lenses (low
completeness) or contaminate the results with non-lenses (low purity).

The first tranche of survey data, introduced from May 2013, is an area
of $\sim172$ square degrees (or a quarter of a percent of the sky)
from the Canada-France-Hawaii Telescope Legacy Survey.  The survey is
divided into patches of $440\times 440$ pixels, each of which is seen
by ten volunteers.  These data had previously been analyzed by robots
\cite{Gavazzi2012, More2012ApJ}.

These encouraging results now raise the question: could modeling of
the lenses also be done by volunteers?  Modeling is a much more
difficult task than searching for lens candidates, as it requires some
expert knowledge.  Nonetheless, a panel of the {\em Spacewarps\/}
volunteers is quite experienced from earlier projects, having spent a
thousand hours or more with data.  Some of these experienced
volunteers are very interested in more demanding projects.  This also
is a general trend \citep[cf.][]{Khatib22112011}.  The present project
then suggests itself.

>>>>

Since those first discoveries, more than 400 secure lenses are now
known (see Figure \ref{fig:masterlens}).  Modelling of the mass
distribution is part of any research using lenses, but so far no
modelling study has spanned all known lenses.  The largest single one
\citep{2009ApJ...703L..51K} models 58 lenses to infer the distribution
of dark matter around galaxies.  In other work,
\cite{2011ApJ...740...97L} combined lens models of 21 galaxies with
models of their stellar populations, to find the relation between
stars and dark matter, and \cite{2014MNRAS.437..600S} modelled 18
time-delay lenses together to infer cosmological parameters.

\begin{figure}
\centering
\includegraphics[width=\hsize]{fig/lenssky}
\caption{Sky distribution of 423 published lenses considered secure
  (from the Masterlens catalog at the University of Utah, maintained
  by Joel Brownstein and Leonidas Moustakas).  The map is in
  Hammer-Aitoff projection, with North up and $\rm RA=0$ in the
  middle. The empty swathes to the left and right of centre are the
  Milky Way. {\em Should we drop this figure?  It's not relevant to
    modelling, but as there seems to be no publicly available summary
    of all known lenses, it does help set context.}}
\label{fig:masterlens}
\end{figure}

Surveys now under way aim to increase the inventory of lenses another
ten or a hundred fold.  Among these is \sw (Marshall et al, in prep,
More et al, in prep).  \sw is a citizen science project\footnote{\tt
  http://www.spacewarps.org} in which volunteers are presented
sky-survey images and invited to identify lens candidates.  Simulated
lenses are mixed in with the data, both to help train volunteers on
what to look for, and to provide measures of the effectiveness of the
search.  The motivation for \sw is to enable volunteers, some of whom
had previously serendipitously identified lens candidates on earlier
citizen-science surveys, to make discoveries missed in automatic
searches by software robots.  Robots are very good with clean lensing
system in uncrowded fields with high signal-to-noise, but in general
test situations \citep{2009ApJ...694..924M} robots miss lenses (low
completeness) or contaminate the results with non-lenses (low purity).
\sw introduced its first tranche of survey data, in May 2013.  The
Canada-France-Hawaii Telescope Legacy Survey (CFHTLS), covering
$\simeq172$ square degrees (0.4\% of the sky) was divided into tiles
of $80''\times80''$ ($440\times 440$ pixels).  Each such tile was
shown to ten volunteers.  Despite the rarity of detectable lenses
($<10^{-3}$ per tile), the \sw collective soon found lens candidates,
both rediscoveries already known from robots \citep{Gavazzi2012,
  More2012ApJ} and possible new lenses.

The encouraging early results prompted the question: could modeling of
the lenses also be done by volunteers?  There are several software
packages available for lens modelling \citep[for a review,
  see][]{2013NewAR..57....1L} and work has been done on generic
interfaces \citep{2014A&C.....5...28L}, lens modelling has remained
the domain of specialists.  An early proposal for what became \sw had
included a prototype lens modelling tool \citep{2010AAS...21543527N},
but the learning curve for modelling was subsequently considered too
long to attract potential volunteers.  On the other hand, some {\em
  Spacewarps\/} volunteers were quite experienced from earlier
projects, having spent a thousand hours or more with data, and very
interested in more demanding projects.  The interests of
citizen-science communities are just beginning to be studied
\citep[e.g.,][]{2013AEdRv..12a0106J}, but it is clear that some
volunteers welcome open-ended challenges, and sometimes these have led
to new scientific results: one example is the discovery of an
exceptional extra-solar planet \citep{2013ApJ...768..127S}; another is
the development of new algorithms for protein folding
\citep{Khatib22112011}.

A characteristic of strong gravitational lensing is that lensed images
do not look much like the source, and still less do they resemble the
mass distribution.  Modellers develop an intuition for image
configurations, and it is necessary to do so before one can begin to
model lenses.  Indeed, the modelling section of
\cite{1981ApJ...244..723Y} can be read as researchers developing
lens-modellers' intuition for the first time.
